Engineering momentum profiles of cold-atom beams

### 2018-08-30 Paper/version1/paper.tex
Added Version 2

\subsection{Determining the Link Potential}
We now present a procedure for determining a link potential that produces a generic desirable flux properties. This simple procedure performs a global search over a family of possible link potentials in order to minimize the difference between the desired flux properties and the actual flux properties. Due to the non-convex nature of this optimization problem, we perform this search using the well-known global optimization routine called Differential Evolution (DE) \cite{original DE paper}. Despite its astounding simplicity, this algorithm does an excellent job of balancing the need to search the entire space against the need to learn from subsequent trials in the optimization process in order to speed up convergence.

>>>>

\subsection{Finding a Link Potential}
Our simple process for finding an appropriate link potential consists of a global search over a family of possible link potentials in order to minimize the difference (more precisly the ``cost'', defined in Eq.~\eqref{eq:cost} below) between the desired and the actual flux properties. We perform this search using the well-known global optimization routine called Differential Evolution (DE) \cite{original DE paper}.This evolutionary-based search is suitable given the non-convex nature of the optimization problem. Despite its simplicity, DE does a good job of balancing the need to search the entire space against the need to learn from subsequent trials in the optimization process in order to speed up convergence with minimal complication in terms of choosing the right model settings. Emperically, we found DE to perform much better than several other approaches including random search, Nelder-Mead, and Simulated Annealing \cite{Tests performed using mathematica}.

Before detailing our procedure, we emphasize that our goal is not to the find the ``best'' link potential in terms of a theoretical cost function; one that is ``good enough'' in terms of experimental requirements will do. This relaxed goal removes many complications. For instance, as long as a discovered potential is {\it good enough} we do not even need to worry about whether or not the DE routine has converged. Similarly, if we run the routine twice and arrive at two very different but similarly {\it good enough} solutions, we do not have to worry about which solution is ``right''. We can just pick our favorite and proceed.

### 2018-08-30 Paper/version2/paper.tex
Added Version 2

\documentclass[onecolumn,amsmath,amssymb,nofootinbib,prl]{revtex4-1}

\usepackage{comment}
\usepackage{graphicx} 
\usepackage{xcolor}
\usepackage{bm}

\usepackage[%frenchlinks=true,
colorlinks=true,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}]{hyperref}

@@

@@

>>>>

\documentclass[twocolumn,amsmath,amssymb,showpacs,prl,superscriptaddress,aps]{revtex4-1}

\usepackage{epsfig}
\usepackage{color}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{epstopdf}

@@

\date{\today}

@@

\begin{acknowledgments}
We thank Michael Fleischhauer for referring to~\cite{paper}.
A.~G.~V. gratefully acknowledges the support of the Humboldt Foundation.
\end{acknowledgments}

\widetext

\section{Supplementary Material}

\section{Polaron}

To model one impurity atom that moves through the one-dimensional environment made of $N$ cold bosonic atoms, we employ the following Hamiltonian
\begin{equation}
H=-\frac{\hbar^2}{2m}\sum_{i=1}^N\frac{\partial^2}{\partial x_i^2}-\frac{\hbar^2}{2M}\frac{\partial^2}{\partial y^2}+g\sum_{i>j=1}^N\delta(x_i-x_j)+c\sum_{i=1}^N \delta(x_i-y),
\end{equation}
where $M$ is the mass of the impurity atom, and $m$ is the mass of a bosonic particle. The position of the impurity is $y$, bosons are described by the set of coordinates $\{x_i\}$. 
We assume that the realistic boson-boson and boson-impurity interactions are well-described by the zero-range potentials of strengths $g$ and $c$ respectively. 
The reservoir by assumption is large, such that the dynamics can be described by the thermodynamic limit $N\to \infty$ assuming a certain value of the density $\rho$.
To take the thermodynamic limit we set the periodic boundary conditions: The particles move in a ring of the circumference $L$, such that $0<x_i<L$ and $0<y<L$.
This assumption is not essential for our problem, because we are interested in the limit $N(L)\to \infty$ with $\rho=\frac{N}{L}$.

Every eigenfunction of the Hamiltonian can be written as $\Psi=\sum_{\{n_j\},m} a_{\{n_j\},m}e^{2\pi i \frac{\sum n_jx_j+my}{L}}$. Because all interactions are pairwise, 
the total (angular) momentum of the system must be conserved, and we write it as $P=\frac{2\pi\hbar}{L}\left(\sum_{j} n_j+m\right)$.
Since we have a conserved quantity ($P$), we may exclude the coordinate of the impurity from the consideration. To this end, 
we write the function $\Psi$ as $\Psi=e^{i \frac{P y}{\hbar}}\sum_{\{n_j\},m} a_{\{n_j\},m}e^{2\pi i \frac{\sum n_jz_j}{L}}\equiv e^{i \frac{P y}{\hbar}} \psi(z_1,...,z_N)$ 
with $z_i=L\theta(x_i-y)+x_i-y$, where $\theta(x)$ is the Heavyside step function, i.e., 
$\theta(x>1) = 1$ and zero otherwise. The variables $z_i$ are defined such that $0\leq z_i \leq L$ and the impurity is placed at $z=0$ and $z=L$. Now if we insert this 
function into the Schr{\"o}dinger equation, $H\Psi={\cal E}\Psi$, we obtain the following equation for $\psi(0<z_i<L)$
\begin{equation}
-\frac{\hbar^2}{2m}\sum_i\frac{\partial^2 \psi}{\partial z_i^2}-\frac{\hbar^2}{2M}\left(\sum_{i}\frac{\partial \psi}{\partial z_i}\right)^2
+ i \frac{\hbar P}{M}\sum_{i}\frac{\partial \psi}{\partial z_i}+g\sum_{i>j}\delta(z_i-z_j)\psi=\left({\cal E}-\frac{P^2}{2M}\right)\psi,
\end{equation}
which must be supplemented with the boundary conditions:
\begin{equation}
\psi(z_i=0)=\psi(z_i=L); \qquad \frac{\partial \psi}{\partial z_i}\bigg|^{z_i=0^+}_{z_i=L^-}= \frac{2 c \kappa}{\hbar^2} \psi(z_i=0),
\end{equation}
where $\kappa=mM/(m+M)$ is the reduced mass.

By assumption the bosons are weakly-interacting, therefore we use the ansatz $\psi=\prod_i \Phi(z_i)$ to approximate the function. 
To minimize the expectation value of the Hamiltonian the function $\Phi(z)$ must satisfy the following Gross-Pitaevski-type equation
\begin{equation}
-\frac{\hbar^2}{2\kappa}\frac{\partial^2\Phi}{\partial z^2}+i\frac{\hbar P}{M}\frac{\partial \Phi}{\partial z}
-i \frac{\hbar^2 (N-1) A}{M} \frac{\partial\Phi}{\partial z} + g(N-1)|\Phi|^2\Phi=\mu \Phi,
\end{equation}
where $A=-i\int \Phi(x)^*\frac{\partial}{\partial x}\Phi(x)\mathrm{d}x$ defines the momentum of a boson, and $\mu$ is the chemical potential (it appears as the Lagrange multiplier). 
We rewrite this equation as
\begin{equation}
-\frac{\partial^2\Phi}{\partial z^2}+i v \frac{\partial \Phi}{\partial z} + \tilde g(N-1)|\Phi|^2\Phi+\tilde c\delta(z)\Phi=\tilde\mu\Phi,
\label{eq:GPE_resc}
\end{equation}
where $\tilde\mu=\frac{2 \kappa \mu}{\hbar^2}$, $\tilde g=\frac{2 \kappa g}{\hbar^2}$, $\tilde c=\frac{2 \kappa c}{\hbar^2}$ and 
$v\equiv \frac{2 \kappa P_I}{M \hbar}$, where $P_I=P-\hbar A(N-1)$ defines the momentum of the impurity in the thermodynamic limit.
Note the corresponding boundary conditions 
\begin{equation}
\Phi(z=0)=\Phi(z=L); \qquad \frac{\partial \Phi}{\partial z}\bigg|^{z=0^+}_{z=L^-}= \tilde c \Phi(0),
\end{equation}

This non-linear equation~(\ref{eq:GPE_resc}) has an analytic solution for the soliton-like behavior~\cite{ishikawa1980,hakim1997}, which determines the properties 
of the polaron in our problem. Let us first consider the case with $c=0$. In this case the solution for $v>0$ is
\begin{equation}
\Phi=\sqrt{\frac{\tilde \mu}{\tilde g(N-1)}}\left(1-\beta \mathrm{sech}^2\left[\sqrt{\frac{\tilde\mu\beta}{2}}(z+z_0)\right]\right)^{\frac{1}{2}}e^{i\phi(z)},
\end{equation}
\begin{equation}
\phi(z)=-\pi\theta(z+z_0)+\mathrm{arctan}\left(\frac{\sqrt{\frac{2 v^2}{\tilde \mu}\beta}}{\mathrm{exp}\left[\sqrt{2\tilde \mu\beta}(z+z_0)\right]-2\beta+1}\right),
\end{equation}
where  $\beta=1- v^2/(2\tilde \mu)$, and $z_0$ is some parameter that should be determined from the boundary conditions. It is worthwhile noting that the solution for $v<0$ is $\Phi^*$. We plot this solution in Figs.~\ref{fig:Fig1} and~\ref{fig:Fig2}; for simplicity we consider the region $-L/2<z<L/2$, the region $0<z<L$ easily follows. Note that the solution is not periodic (see $\phi$), however by combining two solution with $\pm z_0$ 
we can construct a periodic solution with a singularity at $z=0$. Therefore, the polaron in this picture is a combination of two moving solitons: The impurity creates a topological defect, which 
leads to a quasiparticle picture.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centerline{\includegraphics[scale=0.5]{density.pdf}}
\caption{The density, $|\Phi|^2$, of the Bose gas for two different parameters $z_0$: $z_0=0.2$ (left curve), $z_0=-0.2$, assuming that $\tilde \mu=1.6, v=0.3$ (everything in arbitrary units). 
Note that the minimum of the density is at $-z_0$.}
\label{fig:Fig1}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centerline{\includegraphics[scale=0.5]{phase.pdf}}
\caption{The phase, $\phi$, of the Bose gas for two different parameters $z_0$: $z_0=0.2$ (left curve), $z_0=-0.2$, assuming that $\tilde \mu=1.6, v=0.3$ (everything in arbitrary units).}
\label{fig:Fig2}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We write the wave function for the polaron as
\begin{equation}
\Phi=\sqrt{\frac{\tilde \mu}{\tilde g(N-1)}}\left(1-\beta \mathrm{sech}^2\left[\sqrt{\frac{\tilde\mu\beta}{2}}(z\pm z_0)\right]\right)^{\frac{1}{2}}e^{i\phi(z)},
\end{equation}
with 
\begin{equation}
\phi(z)=\delta \phi \theta(-z)+\mathrm{arctan}\left(\frac{\sqrt{\frac{2 v^2}{\tilde \mu}\beta}}{\mathrm{exp}\left[\sqrt{2\tilde \mu\beta}(z\pm z_0)\right]-2\beta+1}\right),
\end{equation}
where the parameter $\delta \phi$ is not important for the further derivations, it reassures that the function is periodic  
(in the thermodynamic limit it is written as $2\pi - \mathrm{arctan}\left[\frac{\sqrt{\frac{2 v^2}{\tilde \mu}\beta}}{1-2\beta}\right]$);
the plus sign in $\pm$ corresponds to $z>0$ and the minus sign to $z<0$. The wave function can be easily visualised from Figs.~\ref{fig:Fig1} and~\ref{fig:Fig2}. 
The density has a non-analytic derivative at $z=0$. The phase is discontinuous function at $z=0$ (the derivative is continuous).

The corresponding chemical potential is found from the normalization condition $\int \Phi^2=1$, for large values of $N$ and $L$ it reads
\begin{equation}
\tilde \mu=\gamma \rho^2 \frac{N-1}{N}\left(1-2\sqrt{2\beta} \frac{ (\mathrm{tanh}(d)-1)}{\sqrt{\gamma}N}\right),
\end{equation}
where $\rho=N/L$, $\gamma=\tilde g/\rho$, and $d=\sqrt{\frac{\gamma \beta}{2}}\rho z_0$.
The condition on $z_0$ is found by using the boundary conditions at $z=\{0,L\}$
\begin{equation}
\frac{\tilde c}{\rho \sqrt{2\gamma}}=\frac{\beta^{\frac{3}{2}}\tanh(d)}{-\beta+\cosh^2(d)}.
\end{equation}
This equation is cubic (in $\tanh(d)$), hence, the solutions can be found in a closed form.
Now we can calculate the energy of the polaron in the thermodynamic limit
\begin{equation}
E\equiv \lim_{N\to\infty, \frac{N}{L}\to \rho}\left[{\cal E}(c,P)-{\cal E}(c=0,P=0)\right],
\end{equation}
where
\begin{equation}
{\cal E}(c,P) = \frac{P^2}{2M} + \mu N-\frac{\hbar^2 A^2 N(N-1)}{2M}-gN(N-1)\int_{0}^{L/2}|\Phi|^4\mathrm{d}z.
\end{equation}
Using these expressions we derive 
\begin{equation}
E=\frac{P_I^2}{2M}+\frac{\hbar^2\rho}{2\kappa}\frac{\sqrt{2 \tilde g \rho \beta}}{3}\left[4 b + (-4b+\beta\mathrm{sech}^2(d))\tanh(d)\right]+\frac{\hbar P_I}{M}\lim_{N\to\infty} A N,
\end{equation}
where $b=1+\frac{v^2}{4\tilde g\rho}=1+\frac{\kappa P_I^2}{2M^2 g\rho}$. This energy for $v\to 0$ can be written as 
\begin{equation}
E\simeq \epsilon+ \frac{P_I^2}{2m_{\mathrm{\mathrm{eff}}}},
\end{equation}
where $\epsilon$ is the effective energy of the polaron, and $m_{\mathrm{eff}}$ is the effective mass.

Further Discussions in the supplementary material and in the paper:

Discuss the effective mass.

Discuss the two different solutions.

Discuss the critical momentum.

Discuss the quench dynamics to the polaron.

Discuss the literature.
